---
title: Close matches
author: James Baird
---

```{r}
#| label: setup
#| include: false

library(here)
library(dplyr)
library(gt)
library(tidyr)
source(here("R", "model.R"))

db <- Model$new()
```

# Close wins

A close win is defined as:

- Winning by 2 or fewer wickets.
- Winning by 20 or fewer runs.

```{r}
#| label: close-matches

wicket_threshold <- 2
run_threshold <- 20

close_matches <-
  db$table("Match") |>
  filter(
    WinWickets <= wicket_threshold | WinRuns <= run_threshold,
    !WinByInnings
  )
```

```{r}
#| label: recent-close-wins

winner_style <-
  list(
    cell_text(weight = "bold"),
    cell_fill(color = "#e6ffe6")
  )

close_matches |>
  transmute(
    StartDate,
    Venue,
    Team1,
    Team2,
    Winner,
    Result = case_when(
      !is.na(WinWickets) ~ paste0(WinWickets, " wickets"),
      !is.na(WinRuns) ~ paste0(WinRuns, " runs")
    )
  ) |>
  arrange(desc(StartDate)) |>
  gt() |>
  tab_style(
    style = winner_style,
    locations = cells_body(
      columns = Team1,
      rows = Winner == Team1
    )
  ) |>
  tab_style(
    style = winner_style,
    locations = cells_body(
      columns = Team2,
      rows = Winner == Team2
    )
  ) |>
  fmt_date(columns = StartDate, date_style = 7) |>
  # Hide winner column
  cols_hide(columns = Winner) |>
  tab_header(
    title = "Most recent close wins",
    subtitle = "Matches won by 2 or fewer wickets, or 20 or fewer runs."
  )
```

```{r}
#| label: close-matches-by-team

close_matches |>
  select(MatchId, Team1, Team2, Winner) |>
  pivot_longer(
    c(Team1, Team2),
    names_to = "index",
    values_to = "team"
  ) |>
  mutate(won = Winner == team) |>
  group_by(team) |>
  summarise(
    n_matches = n(),
    n_wins = sum(won),
    win_pct = mean(won)
  ) |>
  arrange(desc(win_pct)) |>
  collect() |>
  gt() |>
  fmt_percent(columns = win_pct, decimals = 1) |>
  data_color(
    columns = win_pct,
    palette = "Greens"
  ) |>
  cols_label(
    team = "Team",
    n_matches = "Close Matches",
    n_wins = "Wins",
    win_pct = "Win %"
  ) |>
  tab_header(
    title = "Team performance in close matches",
    subtitle = "Matches won by 2 or fewer wickets, or 20 or fewer runs."
  )
```

# Close draws

A close draw is defined as:

- Team being 8 or more wickets down at the end of the 4th innings.
- Team being 20 or fewer runs behind at the end of the 4th innings.

Note that retired hurt dismissals are counted as wickets, as it is
assumed that the player would not have returned, and therefore the
match would have ended.

Some matches could end up in both categories.

```{r}
draws <-
  db$table("Match") |>
  filter(Result == "draw") |>
  select(MatchId)

# Number of runs per innings
innings_runs <-
  db$table("Delivery") |>
  summarise(
    innings_runs = sum(TotalRuns),
    .by = c(MatchId, InningsNumber)
  )

# Total team runs per match
team_runs <-
  innings_runs |>
  inner_join(db$table("Innings"), c("MatchId", "InningsNumber")) |>
  summarise(
    team_runs = sum(innings_runs),
    .by = c(MatchId, BattingTeam)
  )

# Run difference between teams per match
run_difference <-
  team_runs |>
  mutate(team_id = row_number(), .by = MatchId) |>
  select(-BattingTeam) |>
  pivot_wider(
    names_from = team_id,
    values_from = team_runs,
    names_prefix = "team_"
  ) |>
  mutate(
    run_diff = abs(team_1 - team_2)
  )
```

## At least 8 wickets down

```{r}
# Identify 8 wickets in 4th innings
close_draw_wickets <-
  db$table("Wicket") |>
  semi_join(draws, "MatchId") |>
  filter(InningsNumber == 4) |>
  count(MatchId) |>
  filter(n >= 8) |>
  select(MatchId, wickets = n)

close_draw_wickets |>
  left_join(run_difference, "MatchId") |>
  inner_join(db$table("Match"), "MatchId") |>
  collect() |>
  arrange(desc(wickets), run_diff, desc(StartDate)) |>
  select(StartDate, Venue, Team1, Team2, wickets, run_diff) |>
  gt() |>
  fmt_date(columns = StartDate, date_style = 7) |>
  cols_label(
    wickets = "Wickets",
    run_diff = "Run Difference",
    StartDate = "Date",
    Team1 = "Team 1",
    Team2 = "Team 2"
  )
```

## Fewer than 20 runs behind

```{r}
# The match must have been in the 4th innings
four_innings_matches <- db$table("Innings") |>
  filter(InningsNumber == 4) |>
  select(MatchId)

close_draw_runs <-
  run_difference |>
  semi_join(draws, "MatchId") |>
  semi_join(four_innings_matches, "MatchId") |>
  filter(run_diff <= 20) |>
  select(MatchId, run_diff) |>
  inner_join(db$table("Match"), "MatchId") |>
  select(
    StartDate,
    Venue,
    Team1,
    Team2,
    run_diff
  ) |>
  collect()

close_draw_runs |>
  arrange(run_diff, desc(StartDate)) |>
  gt() |>
  fmt_date(columns = StartDate, date_style = 7) |>
  cols_label(
    run_diff = "Run Difference",
    StartDate = "Date",
    Team1 = "Team 1",
    Team2 = "Team 2"
  )
```