---
title: Multi-wicket overs
author: James Baird
---

A multi-wicket over is one in which two or more wickets are taken.

```{r}
#| label: setup
#| include: false
library(here)
library(dplyr)
library(gt)
library(ggplot2)
source(here("R", "model.R"))
source(here("R", "plot.R"))

db <- Model$new()
```

# Frequency

```{r}
wickets_per_over <-
  db$table("Wicket") |>
  group_by(MatchId, InningsNumber, Over) |>
  summarise(n_wickets = n(), .groups = "drop")

all_overs <-
  db$table("Delivery") |>
  distinct(MatchId, InningsNumber, Over) |>
  left_join(wickets_per_over, join_by(MatchId, InningsNumber, Over)) |>
  mutate(n_wickets = coalesce(n_wickets, 0))

multi_prop <-
  all_overs |>
  summarise(multi_prop = mean(n_wickets > 1)) |>
  pull(multi_prop)

all_overs |>
  count(n_wickets) |>
  collect() |>
  mutate(prop = n / sum(n)) |>
  gt() |>
  fmt_integer(n) |>
  fmt_percent(prop) |>
  cols_merge_n_pct(n, prop) |>
  cols_label(n_wickets = "Wickets in over", prop = "N (%) of overs")
```

Multiple wickets in an over is a rare event, occurring in only
`r signif(multi_prop * 100, 2)`% of overs in the database.

However, given that a typical match has hundreds of overs, the occurrence
of at least one multi-wicket over in a match is quite common. 

```{r}
n_multi_wickets <-
  all_overs |>
    group_by(MatchId) |>
    summarise(
      n_multi = sum(n_wickets > 1),
      .groups = "drop"
    ) |>
    count(n_multi) |>
    collect()

n_multi_wickets |>
  ggplot(aes(x = n_multi, y = n)) +
  geom_col(fill = "#004C97") +
  labs(
    x = "Number of multi-wicket overs",
    y = "Number of matches",
    title = "Number of multi-wicket overs in match"
  ) +
  scale_x_continuous(breaks = ~ seq(0, .x[2], 1)) +
  theme_cricket()
```

# Matches with the most multi-wicket overs

```{r}
#| label: matches
wickets_per_over |>
  filter(n_wickets > 1) |>
  count(MatchId) |>
  left_join(
    db$table("Match"),
    "MatchId"
  ) |>
  select(
    StartDate,
    Team1,
    Team2,
    Venue,
    n
  ) |>
  filter(n >= 3) |>
  arrange(desc(n), desc(StartDate)) |>
  collect() |>
  gt() |>
  fmt_integer(n) |>
  fmt_date(
    StartDate,
    date_style = "day_month_year" 
  ) |>
  cols_label(
    Venue = "Venue",
    StartDate = "Date",
    Team1 = "Team 1",
    Team2 = "Team 2",
    n = "Number of overs with multiple wickets"
  ) |>
  opt_interactive()
```

# Bowlers with the most multi-wicket overs

Whereas the previous figures included all wickets in an over,
the following figures only include wickets credited to the bowler.

```{r}
ineligible_methods <- c(
  "retired hurt",
  "retired not out",
  "run out"
)

eligible_wickets <-
  db$table("Wicket") |>
  filter(!Kind %in% ineligible_methods) |>
  group_by(MatchId, InningsNumber, Over) |>
  filter(n() > 1) |>
  ungroup()

# Get distinct overs per bowler
bowler_wicket_overs <-
  eligible_wickets |>
    group_by(MatchId, InningsNumber, Over, Bowler) |>
    summarise(n_wickets_in_over = n(), .groups = "drop")

# Count total number of overs bowled per bowler
all_bowler_overs <-
  db$table("Delivery") |>
    distinct(MatchId, InningsNumber, Over, Bowler) |>
    left_join(
      bowler_wicket_overs,
      join_by(MatchId, InningsNumber, Over, Bowler)
    ) |>
    mutate(n_wickets_in_over = coalesce(n_wickets_in_over, 0))


multi_bowler_summary <-
  all_bowler_overs |>
    group_by(Bowler) |>
    summarise(
      n_overs = n(),
      n_multi = sum(n_wickets_in_over > 1)
    ) |>
    mutate(multi_prop = n_multi / as.double(n_overs)) |>
    filter(n_multi > 1) |>
    left_join(
      db$table("Player"),
      join_by(Bowler == PlayerId)
    ) |>
    select(Name, n_overs, n_multi, multi_prop) |>
    arrange(desc(n_multi)) |>
    collect()
  
multi_bowler_summary |> 
  gt() |>
  fmt_integer(n_overs, n_multi) |>
  fmt_percent(multi_prop) |>
  data_color(
    c(multi_prop, n_multi),
    palette = "Greens"
  ) |>
  cols_label(
    Name = "Bowler",
    n_overs = "Overs bowled",
    n_multi = "Overs with multiple wickets",
    multi_prop = "Proportion of overs with multiple wickets"
  ) |>
  tab_header(
    title = "Bowlers with the most overs with multiple wickets",
    subtitle = "Includes bowlers with at least two overs with multiple wickets"
  ) |>
  opt_interactive()
```