---
title: Wicket with last ball of career
author: James Baird
---

```{r}
#| label: setup
#| include: false

library(here)
library(dplyr)
library(dbplyr)
library(gt)
source(here("R", "model.R"))
source(here("R", "util.R"))

db <- Model$new()
```

This analysis looks for players who took a wicket with
the last ball of their career. If the player did not bowl
in the final match of their career, they will be included
based on the match where they bowled their final ball.
Bowlers will commonly appear here if they took the 10th
wicket in the 3rd or 4th innings of the most recent Test
their team played.

This analysis assumes that players will not play another match, thus players who are still playing may transiently
appear in this list until they bowl in another match.

```{r}
#| label: data-prep

# Index each bowler's deliveries and get last ball per bowler
last_ball <-
  db$table("Delivery") |>
  inner_join(db$table("Match"), join_by(MatchId)) |>
  group_by(Bowler) |>
  window_order(StartDate, InningsNumber, Over, Ball) |>
  mutate(
    delivery_index = row_number()
  ) |>
  slice_max(delivery_index, n = 1) |>
  ungroup() |>
  select(Bowler, MatchId, InningsNumber, Over, Ball) |>
  compute()

# Get wickets (excluding non-bowler dismissals)
wickets <-
  db$table("Wicket") |>
  filter(!Kind %in% !!get_non_bowler_dismissals()) |>
  select(Bowler, InningsNumber, Over, Ball)

last_ball |>
  inner_join(wickets, join_by(Bowler, InningsNumber, Over, Ball)) |>
  inner_join(db$table("Match"), join_by(MatchId)) |>
  inner_join(db$table("Player"), join_by(Bowler == PlayerId)) |>
  select(StartDate, Name) |>
  arrange(StartDate) |>
  collect() |>
  gt() |>
  common_label() |>
  fmt_date(StartDate, date_style = 7) |>
  tab_header(
    title = "Wicket with last ball of career"
  )
```